Semantic Authoring in a Flexiformal Context --- Bulk Annotation of Rigorous Documents

### 2025-07-25 doc/cicm25-paper/paper.tex
finalized paper?

\documentclass[runningheads]{llncs}
\usepackage[show]{ed}
\usepackage{wrapfig}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{stex-logo}
\usepackage{paralist}
\usepackage[hidelinks]{hyperref}
\usepackage{orcidlink}
%\usepackage{lststex} debug that the gray is not solid
\usepackage{listings}
\definecolor{codegray}{rgb}{0.9,0.9,0.9}
\lstset{basicstyle=\sf,columns=fullflexible,backgroundcolor = \color{codegray}}
\lstset{numberstyle=\tiny}
\lstset{language={[LaTeX]TeX}}
\usepackage[style=alphabetic,hyperref=auto,defernumbers=true,backend=bibtex,firstinits=true,maxbibnames=9,maxcitenames=3,isbn=false]{biblatex}
\addbibresource{kwarcpubs.bib}
\addbibresource{extpubs.bib}
\addbibresource{kwarccrossrefs.bib}
\addbibresource{extcrossrefs.bib}
\addbibresource{otherrefs.bib}
\usepackage[noabbrev]{cleveref}
\usepackage{tikz}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{docicon}
\usepackage{soul}
\def\edited#1{\hl{#1}}

@@

\newcommand\ALeA{\textsf{ALeA}\xspace}
\newcommand\SMGloM{\textsf{SMGloM}\xspace}
\newcommand\snify{\textsf{snify}\xspace}
\newcommand\defianno{\textsf{defianno}\xspace}
\newcommand\FTML{\textsf{FTML}\xspace}
\newcommand\WOIDE{\textsf{WOIDE}\xspace}
\def\llangle{\langle\kern-.2em\langle}
\def\rrangle{\rangle\kern-.2em\rangle}
\def\flams{\textsf{FLAMS}\xspace}

@@

% \ednote{
%     \textbf{Related work: semantic web authoring}
%     \emph{semantic authoring} already used in context of semantic web (RDFa).
%     Tools exist and it doesn't have exactly the same challenges as we face, though there is some overlap.
%     We have a module system and need more dependency management, something else?
%     This has to be described somewhere.
% 
%     Related work: ``User Interfaces for Semantic Authoring of Textual Content: A Systematic Literature Review.''
% }

@@

\begin{figure}[h]\centering
  \fbox{\includegraphics[width=10cm,trim=0.8cm 0 0 0]{../img/minimax-remarks.pdf}}
  \caption{An unannotated learning object from an AI lecture}\label{fig:lo}
\end{figure}

\subsection{Running Example: Annotating ``terminal''}
With \sTeX, authors can add semantic references to technical terms,
which enables straightforward learning support services like
showing the definition of a term, but also informs the system
about the content/prerequisites of a learning object.
Adding these references is a key practical problem when converting presentational {\LaTeX}
course materials into semantic documents.
\begin{figure}[h]\centering
  \fbox{\includegraphics[width=12cm]{../img/search-problem.en.pdf}}
\begin{lstlisting}[morekeywords={definame,symdecl,definiendum},numbers=left,
escapechar=!]
\begin{smodule}[title=Search Problem]{search-problem}
[... some imports ...]
\symdecl*{goal state}
\begin{sdefinition}
  [...] Certain \sns{state} are [...] \definame[post=s]{goal state} [...]
  (also called \definiendum{goal state}{terminal states}).
\end{sdefinition}
\end{smodule}
\end{lstlisting}
  \caption{Simplified definition of ``goal state'' from the domain model}\label{fig:state-space}
\end{figure}

Take for instance a learning object like the paragraph in \Cref{fig:lo}\ednote{MK: in
  diesem Bild sollten keine blauen/annotierten Wörter sein}, where we want to annotate the
term ``terminal'' with the (semantic) concept of ``a goal state of a search problem''. The
latter can either be defined earlier in the course or in the domain model (or
both). Concretely this is about converting the unannotated {\LaTeX} string

\begin{lstlisting}[numbers=left,firstnumber=3,
% caption=The unannotated {\LaTeX} sources of \Cref{fig:lo},label=lst:los
]
\blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is terminal, we use its utility,
and otherwise an estimate.
\end{lstlisting}
into the partially annotated 
\begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
% caption=Annotating ``terminal'' in \Cref{lst:los},label=lst:losa
]
\importmodule[smglom/search]{mod?search-problem}
[...]
\blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its utility,
and otherwise an estimate.
\end{lstlisting}
where the author has annotated the word ``terminal'' with the concept it refers to via the
\sTeX macro
\lstinline[mathescape]|\sr{$\llangle \mathit{symbol} \rrangle$}{$\llangle \mathit{verbalization}\rrangle$}|.
In \sTeX, concepts are represented by \textbf{symbols}, which are identified globally and
unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module name}, and a
\textbf{symbol name}.  A \textbf{verbalization} is the natural language rendering of a
concept.  Verbalizations are, of course, language-dependent and can vary between
subdomains and communities of practice, even if the underlying symbol is the same.  In our
example, the verbalization ``terminal [state]'' was annotated with the symbol with name
\lstinline|goal state|. It comes from the module with name \lstinline|search-problem| in
the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
it suffices to use a relative URI in the \lstinline|\sr| annotation. If the symbol name
and the desired verbalization coincide (as they often do in English), the short-hand macro
\lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| for ``utility'').
% in \Cref{lst:losa}).

The annotation refers to a definition in the domain model module shown in
\Cref{fig:state-space}.  The \lstinline|\definame| and \lstinline|\definiendum| in lines
5--6 introduce the verbalizations ``goal state'' and ``terminal state'' for the symbol
\lstinline|search-problem?goal state| introduced by the \lstinline|\symdecl*| declaration in
line 3.

@@

\paragraph{Traditional, informal authoring}
In traditional authoring, e.g. of a scientific article or a textbook, the domain model --
a highly structured model of the domain of discourse -- is in the authors' heads, and is
partially verbalized e.g. in the preliminaries section or the main sections of the
respective article. The content is informal (usually a natural language like English
augmented with technical jargon, formulae, tables, and diagrams) and designed for
processing via the brain of a colleague or student -- in any case, a human who shares a
canonical part of the domain model with the authors. Consequently a large part of the
authoring challenge lies in predicting the state and extent of the domain model of the
reader and creating text that bridges the gap between the reader's domain model and the
payload content of the article -- the knowledge the article or textbook intends to impart
-- ideally using the technical vocabulary the reader is familiar with to reduce this
gap. Tool support for traditional authoring is minimal and usually restricted to
spell/grammar-checkers and online thesauri if we disregard LLM-based writing/formulation
support for now.
\edited{
In our experience, LLM-based authoring support is less helpful in technical documents
than in everyday writing.
}

@@

In addition to the base text, semantic authoring also has to provide the semantic
annotations that drive the personalization and interaction downstream. Techniques from formal
authoring could help here, but in contrast to formal authoring, we cannot just trigger the
IDE functionality by the formal syntax, as the authored content is -- so far -- informal
and the content commons (the target of the annotations) is itself only flexiformal. In a
way, we need a local flexiformalization workflow that helps authors bridge the
informal-formal gap.

@@

Efficient support for the annotation of technical terms in natural language
has been elusive so far.

@@

\paragraph{Evaluation}
% \section{Practical Evaluation and Future Work}\label{sec:eval}
\begin{wrapfigure}{r}{0.4\textwidth}
    \vskip-1.2cm
  \centering
  \includegraphics[width=0.4\textwidth]{../img/annocounts.pdf}
    \vskip-0.2cm
  \caption{Number of annotated terms in the \sTeX/\ALeA commons over time.}\label{fig:annocounts}
  \vskip-0.7cm
\end{wrapfigure}
In our experience, \snify's step-through workflow for annotating term references 
% for symbols from the domain model 
is almost an order of magnitude more efficient than writing the
annotations and imports by hand.\footnote{
        The precise productivity gain depends on the annotator's
        familiarity with the domain model (\edited{searching and identifying}
        the right symbol is time-consuming) but also on other factors like
        the typing speed.
        Measurements are complicated by the fact that
        authors tend to annotate more terms when using \snify.
        We have observed speed-ups over IDE-based annotation
        ranging from 3 (for an expert knowing the domain model by heart)
        to 13 (for a beginner).
        The focus mode further increases productivity
        when considering the overall corpus quality.
}
% , even for annotators who are familiar with the domain model.
Given the sheer number of term annotations (see \Cref{fig:annocounts}),
this results in a considerable productivity gain.
For annotators unfamiliar with the domain model, the unassisted annotation task is
almost infeasible, and the average unfamiliarity naturally grows with the domain
model. The symbol disambiguation process -- on average a word induces 4.1 choices
% \ednote{
%     more details? median is 3, for existing annotations, the average is 2.6 and the median 2...
% }
-- is still manageable,
requires considerable concentration and domain knowledge, but little knowledge of the
domain model flexiformalization.

@@

\begin{wrapfigure}r{5.5cm}\vspace*{-2em}
  \begin{tikzpicture}[scale=.7]
    \tikzstyle{doc}=[draw,thick,align=center,color=black,
                  shape=document,minimum width=10mm,minimum height=15mm]
    \tikzstyle{database}=[cylinder,shape border rotate=90,aspect=0.25,draw, 
     cylinder uses custom fill,cylinder body fill=yellow!30,cylinder end fill=yellow!30]
     \tikzstyle{include}=[right hook-angle 45,thick]
     \tikzstyle{includeleft}=[left hook-angle 45,thick]
    \node[doc] (cc) at (0,-1) {\shortstack{Content\\ Commons}};
    \node[doc] (pl) at (5,-1) {\shortstack{Private/\\Local}};
    \node[database] (db) at (3,2.5) {\shortstack{\tiny Verbalization\\Cache}};
    \node[draw] (sn) at (0,4.5) {\snify};
    \node[draw] (id) at (2.5,5) {IDE};
    \node[draw] (wo) at (5,4.5) {\WOIDE};
    \draw[draw,fill=black!10] (-1.3,.5) rectangle ++ (6.4,1); 
    \node (f) at (-0.3,1.0) {\flams};
      \draw[include]  (cc) -- node[left] {\tiny \begin{tabular}{c}RDF\\exp.\end{tabular}} (db);
    \draw[includeleft] (pl) -- node[right] {\tiny linter} (db);
    \draw[->] (db) -- (sn);
    \draw[->] (db) -- (id);
    \draw[->] (db) -- (wo);
    \draw[dashed,thick] (3,1.6)  -- (3,-2);
    \draw[dashed,thick] (3,1.6)  -- (3,-2);
    \draw[includeleft] (pl) to[bend right=10] (wo);
  \end{tikzpicture}
  \caption{Architecture}\label{fig:arch}\vspace*{-1em}
\end{wrapfigure}

>>>>

\documentclass[runningheads]{llncs}
\usepackage[show]{ed}
\usepackage{wrapfig}
\usepackage{xspace}
\usepackage{graphicx}
\usepackage{stex-logo}
\usepackage{paralist}
\usepackage[hidelinks]{hyperref}
\usepackage{orcidlink}
%\usepackage{lststex} debug that the gray is not solid
\usepackage{listings}
\definecolor{codegray}{rgb}{0.9,0.9,0.9}
\lstset{basicstyle=\sf,columns=fullflexible,backgroundcolor = \color{codegray}}
\lstset{numberstyle=\tiny}
\lstset{language={[LaTeX]TeX}}
\usepackage[style=alphabetic,hyperref=auto,defernumbers=true,backend=bibtex,firstinits=true,maxbibnames=9,maxcitenames=3,isbn=false]{biblatex}
\usepackage{xurl}
\addbibresource{kwarcpubs.bib}
\addbibresource{extpubs.bib}
\addbibresource{kwarccrossrefs.bib}
\addbibresource{extcrossrefs.bib}
\addbibresource{otherrefs.bib}
\usepackage[noabbrev]{cleveref}
\usepackage{tikz}
\usetikzlibrary{shapes}
\usetikzlibrary{arrows}
\usetikzlibrary{docicon}
\usepackage{soul}
% \def\edited#1{\hl{#1}}
\def\edited#1{#1}

@@

\newcommand\ALeA{\textsf{ALeA}\xspace}
\newcommand\SMGloM{\textsf{SMGloM}\xspace}
\newcommand\snify{\textsf{snify}\xspace}
\newcommand\defianno{\textsf{defianno}\xspace}
\newcommand\FTML{\textsf{FTML}\xspace}
\newcommand\WOIDE{\textsf{WOIDE}\xspace}
\def\llangle{\langle\kern-.2em\langle}
\def\rrangle{\rangle\kern-.2em\rangle}
% \def\flams{\textsf{FLAMS}\xspace}

@@

@@

\begin{figure}\centering
  \fbox{\includegraphics[width=10cm,trim=0.8cm 0 0 0]{../img/minimax-remarks.pdf}}
  \caption{An unannotated learning object from an AI lecture}\label{fig:lo}
\end{figure}

\subsection{Running Example: Annotating ``terminal''}
With \sTeX, authors can add semantic references to technical terms,
which enables straightforward learning support services like
showing the definition of a term, but also informs the system
about the content/prerequisites of a learning object.
Adding these references is a key practical problem when converting presentational {\LaTeX}
course materials into semantic documents.

Take for instance a learning object like the paragraph in \Cref{fig:lo},
    where we want to annotate the
term ``terminal'' with the (semantic) concept of ``a goal state of a search problem''. The
latter can either be defined earlier in the course or in the domain model (or
both). Concretely this is about converting the unannotated {\LaTeX} string
\begin{lstlisting}[numbers=left,firstnumber=3,
% caption=The unannotated {\LaTeX} sources of \Cref{fig:lo},label=lst:los
]
\blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is terminal, we use its utility,
and otherwise an estimate.
\end{lstlisting}
into the partially annotated 
\begin{lstlisting}[morekeywords={sr,importmodule},numbers=left,
% caption=Annotating ``terminal'' in \Cref{lst:los},label=lst:losa
]
\importmodule[smglom/search]{mod?search-problem}
[...]
\blue{Note:} Depth-limited minimax requires an evaluation for every
cut-off state $s$. If $s$ is \sr{goal state}{terminal}, we use its utility,
and otherwise an estimate.
\end{lstlisting}
where the author has annotated the word ``terminal'' with the concept it refers to via the
\sTeX macro
\lstinline[mathescape]|\sr{$\llangle \mathit{symbol} \rrangle$}{$\llangle \mathit{verbalization}\rrangle$}|.
In \sTeX, concepts are represented by \textbf{symbols}, which are identified globally and
unambiguously by a URI that includes a \textbf{namespace}, a \textbf{module name}, and a
\textbf{symbol name}.  A \textbf{verbalization} is the natural language rendering of a
concept.  Verbalizations are, of course, language-dependent and can vary between
subdomains and communities of practice, even if the underlying symbol is the same.  In our
example, the verbalization ``terminal [state]'' was annotated with the symbol with name
\lstinline|goal state|. It comes from the module with name \lstinline|search-problem| in
the namespace \lstinline|smglom/search/mod| imported via the \lstinline|\importmodule|
macro in line 1. As the symbol name \lstinline|goal state| is unique in the imports here
it suffices to use a relative URI in the \lstinline|\sr| annotation. If the symbol name
and the desired verbalization coincide (as they often do in English), the short-hand macro
\lstinline|\sn| can be used (e.g. as \lstinline|\sn{utility}| for ``utility'').
% in \Cref{lst:losa}).

The annotation refers to a definition in the domain model module shown in
\Cref{fig:state-space}.  The \lstinline|\definame| and \lstinline|\definiendum| in lines
5--6 introduce the verbalizations ``goal state'' and ``terminal state'' for the
\lstinline|goal state| symbol introduced by the \lstinline|\symdecl*| declaration in
line 3.

\begin{figure}\centering
  \fbox{\includegraphics[width=12cm]{../img/search-problem.en.pdf}}
\begin{lstlisting}[morekeywords={definame,symdecl,definiendum},numbers=left,
escapechar=!]
\begin{smodule}[title=Search Problem]{search-problem}
[... some imports ...]
\symdecl*{goal state}
\begin{sdefinition}
  [...] Certain \sns{state} are [...] \definame[post=s]{goal state} [...]
  (also called \definiendum{goal state}{terminal states}).
\end{sdefinition}
\end{smodule}
\end{lstlisting}
  \caption{Simplified definition of ``goal state'' from the domain model}\label{fig:state-space}
\end{figure}

@@

\paragraph{Traditional, informal authoring}
In traditional authoring, e.g. of a scientific article or a textbook, the domain model --
a highly structured model of the domain of discourse -- is in the authors' heads, and is
partially verbalized e.g. in the preliminaries section or the main sections of the
respective article. The content is informal (usually a natural language like English
augmented with technical jargon, formulae, tables, and diagrams) and designed for
processing via the brain of a colleague or student -- in any case, a human who shares a
canonical part of the domain model with the authors. Consequently a large part of the
authoring challenge lies in predicting the state and extent of the domain model of the
reader and creating text that bridges the gap between the reader's domain model and the
payload content of the article -- the knowledge the article or textbook intends to impart
-- ideally using the technical vocabulary the reader is familiar with to reduce this
gap. Tool support for traditional authoring is minimal and usually restricted to
spell/grammar-checkers and online thesauri if we disregard LLM-based writing/formulation
support for now.
In our experience, LLM-based authoring support is less helpful in technical documents
than in everyday writing.

@@

In addition to the base text, semantic authoring also has to provide the semantic
annotations that drive the personalization and interaction downstream. Techniques from formal
authoring could help here, but in contrast to formal authoring, we cannot just trigger the
IDE functionality by the formal syntax, as the authored content is -- so far -- informal
and the content commons (the target of the annotations) is itself only flexiformal. In a
way, we need a local flexiformalization workflow that helps authors bridge the
informal-formal gap.
We have experimented with LLM-based generation of annotated (\sTeX) content,
but the results have been disappointing so far.

@@

Efficient support for the annotation of technical terms in natural language
has been elusive so far -- we will address this in the next section.

@@

% \newpage
\paragraph{Evaluation}
% \section{Practical Evaluation and Future Work}\label{sec:eval}
In our experience, \snify's step-through workflow for annotating term references 
% for symbols from the domain model 
is almost an order of magnitude more efficient than writing the
annotations and imports by hand.\footnote{
        The precise productivity gain depends on the annotator's
        familiarity with the domain model (\edited{searching and identifying}
        the right symbol is time-consuming) but also on other factors like
        the typing speed.
        Measurements are complicated by the fact that
        authors tend to annotate more terms when using \snify.
        We have observed speed-ups over IDE-based annotation
        ranging from 3 (for an expert knowing the domain model by heart)
        to 13 (for a beginner).
        The focus mode further increases productivity
        when considering the overall corpus quality.
}
% , even for annotators who are familiar with the domain model.
Given the sheer number of term annotations (see \Cref{fig:annocounts}),
this results in a considerable productivity gain.
For annotators unfamiliar with the domain model, the unassisted annotation task is
\begin{wrapfigure}{r}{0.4\textwidth}
    \vskip-0.7cm
  \centering
  \includegraphics[width=0.4\textwidth]{../img/annocounts.pdf}
    % \vskip-0.2cm
  \caption{Number of annotated terms in the \sTeX/\ALeA commons over time.}\label{fig:annocounts}
  \vskip-0.7cm
\end{wrapfigure}
almost infeasible, and the average unfamiliarity naturally grows with the domain
model. The symbol disambiguation process -- on average a word induces 4.1 choices
% \ednote{
%     more details? median is 3, for existing annotations, the average is 2.6 and the median 2...
% }
-- is still manageable,
requires considerable concentration and domain knowledge, but little knowledge of the
domain model flexiformalization.

@@

\begin{wrapfigure}r{5.5cm}\vspace*{-2em}
  \begin{tikzpicture}[scale=.7]
    \tikzstyle{doc}=[draw,thick,align=center,color=black,
                  shape=document,minimum width=10mm,minimum height=15mm]
    \tikzstyle{database}=[cylinder,shape border rotate=90,aspect=0.25,draw, 
     cylinder uses custom fill,cylinder body fill=yellow!30,cylinder end fill=yellow!30]
     \tikzstyle{include}=[right hook-angle 45,thick]
     \tikzstyle{includeleft}=[left hook-angle 45,thick]
    \node[doc] (cc) at (0,-1) {\shortstack{Content\\ Commons}};
    \node[doc] (pl) at (5,-1) {\shortstack{Private/\\Local}};
    \node[database] (db) at (3,2.5) {\shortstack{\tiny Verbalization\\Cache}};
    \node[draw] (sn) at (0,4.5) {\snify};
    \node[draw] (id) at (2.5,5) {IDE};
    \node[draw] (wo) at (5,4.5) {\WOIDE};
    \draw[draw,fill=black!10] (-1.3,.5) rectangle ++ (6.4,1); 
    \node (f) at (-0.3,1.0) {\flams};
      \draw[include]  (cc) -- node[left] {\tiny \begin{tabular}{c}RDF\\exp.\end{tabular}} (db);
    \draw[includeleft] (pl) -- node[right] {\tiny linter} (db);
    \draw[->] (db) -- (sn);
    \draw[->] (db) -- (id);
    \draw[->] (db) -- (wo);
    \draw[dashed,thick] (3,1.6)  -- (3,-2);
    \draw[dashed,thick] (3,1.6)  -- (3,-2);
    \draw[includeleft] (pl) to[bend right=10] (wo);
  \end{tikzpicture}
  \caption{Architecture}\label{fig:arch}\vspace*{-1em}
  \vskip-0.5cm
\end{wrapfigure}
